---
title: "Weight Lifting Exercise Classification Report"
author: "Salman Saleem"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    theme: journal
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE)
```

## 1. Introduction
This report describes the development of a machine learning model to classify weightlifting exercises into five categories (A, B, C, D, E) based on sensor data from accelerometers.

Data Sources
Training Data: pml-training.csv (19,622 observations)

Test Data: pml-testing.csv (20 observations)

## 2 Data Loading and Preprocessing
#2.1 Load Libraries and Data
```{r }
# Load Libraries and Data
library(caret)
library(randomForest)
library(dplyr)
library(tidyr)

# Download data if not already present
if (!file.exists("pml-training.csv")) {
  download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv",
                "pml-training.csv", method = "curl")
}
if (!file.exists("pml-testing.csv")) {
  download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv",
                "pml-testing.csv", method = "curl")
}

# Load data
train_data <- read.csv("pml-training.csv", na.strings = c("NA", "", "#DIV/0!"))
test_data <- read.csv("pml-testing.csv", na.strings = c("NA", "", "#DIV/0!"))
```

## 2.2 Data Cleaning

You can also embed plots, for example:

```{r }
# Remove columns with >95% NA values
train_data <- train_data[, colMeans(is.na(train_data)) < 0.95]
test_data <- test_data[, colMeans(is.na(test_data)) < 0.95]

# Remove metadata columns (non-predictive)
cols_to_remove <- c("X", "user_name", "raw_timestamp_part_1", 
                   "raw_timestamp_part_2", "cvtd_timestamp",
                   "new_window", "num_window")
train_data <- train_data %>% select(-all_of(cols_to_remove))
test_data <- test_data %>% select(-all_of(cols_to_remove))

# Convert classe to factor
train_data$classe <- factor(train_data$classe, levels = c("A", "B", "C", "D", "E"))

# Ensure all remaining columns are numeric
numeric_cols <- sapply(train_data, is.numeric)
train_data <- train_data[, numeric_cols | names(train_data) == "classe"]
test_data <- test_data[, numeric_cols[names(numeric_cols) %in% names(test_data)]]
```
## 3. Data Partitioning
Split the training data into training (70%) and validation (30%) sets

```{r }
set.seed(123)
train_idx <- createDataPartition(train_data$classe, p = 0.7, list = FALSE)
training <- train_data[train_idx, ]
validation <- train_data[-train_idx, ]
```
## 4. Model Training (Random Forest)

``` {r }
# Train model with cross-validation
ctrl <- trainControl(method = "cv", number = 5)
model_rf <- train(classe ~ ., 
                 data = training, 
                 method = "rf",
                 trControl = ctrl,
                 ntree = 100,
                 importance = TRUE)

# View model details
print(model_rf)
```
## 5. Model Evaluation

```{r}
# Validation Set Performance
pred_val <- predict(model_rf, validation)
confusionMatrix(pred_val, validation$classe)
```
## 6. Final Predictions on Test Set
```{r }
test_pred <- predict(model_rf, test_data)
test_pred
```
